No meaningful effects of COVID-19 related social media use on well-being

### 2021-12-01 manuscript.Rmd
- updated language

abstract          : "In times of crisis such as the Corona pandemic, it is important that citizens stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop 'doomscrolling'. In this preregistered study, I investigate whether using social media for COVID-19 related reasons might put personal well-being at risk. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects cross lagged panel models, controlling for several stable and varying covariates. Results showed that the effects of various types of COVID-19 related social media use on well-being were very small, arguably too small to matter. The findings suggest that fears that social media use during times of crisis impairs well-being are likely to be unfounded."
keywords          : "COVID-19, Coronavirus, well-being, affect, life satisfaction, social media use, news use, communication, random effects within between model, panel study, longitudinal."

@@

During the COVID-19 pandemic, 
numerous events unfolded in quick succession.
Several open questions emerged.
How dangerous is the virus? 
Is it spreading in my region? 
How is it transmitted, and how can I protect myself?
Because for many it was (and continues to be) a matter of life or death, people aimed to stay informed regarding the latest developments.
Governments around the world implemented safety measures, such as wearing masks, keeping physical distance, or enforcing lockdowns.
In this extraordinary situation, many people used media excessively to attain information, and especially social media were at an all time high [@statistaAverageDailyTime2021].

Some people could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed "doomscrolling" emerged:
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children [@kleinDarklySoothingCompulsion2021].
As doomscrolling increased it became doubtful whether such social media is helpful, or whether it created an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
These concerns seem justified: 
A study with 6,233 people from Germany found that "[f]requency, duration and diversity of media exposure were positively associated with more symptoms of depression" [@bendauAssociationsCOVID19Related2021, p. 283], which implies negative consequences for well-being.

@@

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated communication [@meierComputermediatedCommunicationSocial2020a].
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
<!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
Well-being can be differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018].
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being, because it represents a general appraisal of one's life.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020].
<!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
To capture well-being in this study, I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a].
Whereas the first four levels focus on the _channel_, the last two address the _type_ of communication.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together provides a nuanced understanding of communication.

First, I investigate how different types of communication affect well-being.
Specifically, I differentiate between active and passive use.
I distinguish (a) _reading_ (passive), _posting_ (active), and _liking and sharing_ COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affect well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially different effects. 
Twitter might have a different effect as compared to WhatsApp because of their different affordances.
For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. 
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about _general_ social media use during times of COVID, but on social media use _focused_ on COVID-19 related content. 
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

@@

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, and found both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
Together, the literature emphasizes potentially negative effects of social media as news use [see also @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
However, note that all of these findings represent between-person relation stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

The question of whether and how _general_ social media use affect well-being, on the other hand, is well-researched.
This also holds true for the different types of communication such as active or passive use.
A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small [@meierComputermediatedCommunicationSocial2020a]---potentially too small to matter.
But what determines whether or not an effect is considered small or trivial?
If we refer to standardized effect sizes, according to @cohenPowerPrimer1992 small effect sizes start at _r_ = .10.
And indeed, several if not most of the current meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @fergusonThisMetaanalysisScreen2021].

These overviews are well aligned with several individual new studies employing advanced methods [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020].
For example, @beyensSocialMediaUse2021 reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the majority the effects were neutral.
In conclusion, most effects are likely somewhere between trivial and small.
I therefore expect that also in the case of COVID-19 related social media use effects will be trivial to small.

@@

First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which is more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news might be depressing, perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects.
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

@@

Together, the strongest argument seems to be that _in general_ the effects of social media on well-being are, on average, small at best.
Because this study only looks at _one part_ of social media use---namely, COVID-19 related interactions---it is very focused, and the overall potential of the effects should diminish further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->
Therefore, I expect that COVID-19 related communication on social media does not affect well-being in a meaningful or relevant way.

@@

Testing this hypothesis, however, is not trivial.
First, in contrast to most hypotheses typically posited in the social sciences, it implicitly contains an effect size, a so-called smallest effect size of interest (SESOI).
Effectively testing this hypothesis necessitates defining what's considered a "trivial effect size" and what's not.
Above I already referred to standardized effect sizes.
However, standardized effect sizes should only be a first step toward evaluating an effect's relevance [@baguleyStandardizedSimpleEffect2009].
Standardized effect sizes are determined by a sample's variance.[^cohensd], which is problematic: 
The question of whether or not social media use affects a particular person in a relevant way should not depend on the variance in the sample in which that person's data were collected.
Instead, it should depend on absolute criteria.

@@

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. 
Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to @normanInterpretationChangesHealthrelated2003, people can reliably distinguish _seven_ levels of satisfaction with health.
So if satisfaction is measured on a 7-point scale, we would state that a four unit change in social media use should result in a one unit change in life satisfaction.

@@

To explain, consider the following example.
Imagine that a person suddenly starts using social media much more than usual, and then after some time  becomes less satisfied with their life.
Eventually, use and life satisfaction return to prior levels.
If this happens to several people at the same time, in a longitudinal study we could then observe a significant effect of social media use on life satisfaction.
However, it could also be the case that during the study there was a major exogenous event (say, a pandemic), which caused large parts of the working population to loose their jobs.
Hence, the causal effect reported above was confounded, because in reality it was the pandemic that caused both social media use to rise and life satisfaction to go down.

Thus, only if we can control for _all_ confounders, can we correctly estimate causality without bias [@rohrerThinkingClearlyCorrelations2018].
Obviously, we can never be entirely sure that we have included all confounders, which makes absolute statements regarding causality virtually impossible.
In addition, when determining the overall causal effect, we need to make sure _not_ to control for mediating variables [@rohrerThinkingClearlyCorrelations2018], for doing so would bias our assessment of the causal effect.
Complicating matters, often it is unclear if a variable is a mediator or a confounder.[^collider]
However, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly.
The aim should therefore be to collect as many varying and nonvarying confounders as possible, while knowing that absolute certainty regarding causality cannot be reached.

@@

When searching for suitable candidates for confounders, we look for variables that affect both media use and well-being. 
Controlling for these factors isolates the actual effect of social media use on well-being.
We can also control for variables that affect only social media use or well-being. 
However, in doing so not much is gained or lost, because the effects of social media use would remain virtually the same [@klinePrinciplesPracticeStructural2016; but see @mcelreathYesterdayClass2021].

In this study, I hence plan to control for the following variables, which either have already been shown to affect both social media use and well-being or which are very likely to do so, and which also are not mediators:
<!-- (I'll additionally include variables that likely affect only well-being, also to obtain a comparison benchmark for social media effects): -->
gender, age, education, Austria country of birth, Austria country of birth of parents, text-based news consumption, video-based news consumption, residency Vienna, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, satisfaction with democracy, disposition to take risks, and locus of control.
I will not control for variables such as trust in institutions or trust in media, because these variables are likely influenced by social media use to a meaningful extent.

@@

Finally, one precondition of causality is temporal order. 
The cause needs to precede the effect.
Finding the right interval between cause and effect is crucial.
For example, if we want to measure the effect of alcohol consumption on driving performance, it makes a big difference if driving performance is measured one minute, one hour, one day, or one week after consumption.
<!-- Finding the right interval is difficult. -->
If variables are stable, longer intervals are needed; if they fluctuate, shorter intervals are necessary.
In the case of well-being, we need shorter intervals for affect and longer ones for life satisfaction.
Still, choosing the right interval is challenging, because especially short intervals are hard to implement in practice and often require advanced methods such as experience sampling (also known as in situ measurement or ambulant assessment) [@schnauber-stockmannMobileDevicesTools2020].

@@

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, p-value) I preregistered on the Open Science Framework. 
The (anonymous) preregistration can be accessed here: https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3.
Because in this study I analyzed data from an already existing large-scale data set, all of these steps were done prior to accessing the data.
The preregistration was designed on the basis of the panel documentation online [@kittelAustrianCoronaPanel2020].
In some cases I could not execute the analyses as I had originally planned, for example because some properties of the variables only became apparent when inspecting the actual data.
The most relevant deviations are reported below, and a complete list of all changes can be found in the online [companion website](https://tdienlin.github.io/Austrian_Corona_Panel/index.html) (https://tdienlin.github.io/Austrian_Corona_Panel/index.html).

@@

Because the data were analyzed post-hoc, no a-priori sample size planning on the basis of power analyses was conducted.
The sample is very large, and it is hence well-equipped reliably to detect also small effects.
<!-- , which is why no exact post hoc power analysis were conducted. -->
In addition, because such large samples easily generate significant _p_-values even for very small effects, it helps that the hypotheses were tested with a smallest effect size of interest-approach.
To this end, I adopted the interval testing approach as proposed by @dienesUsingBayesGet2014.
On the basis of the SESOI, I then defined a null region (in Bayesian terms, this is called the region of practical equivalence, ROPE).
In what follows, I explain how I determined the SESOI and the null region.

In this study, life satisfaction was measured on an 11-point scale.
If people can reliably differentiate 7 levels as mentioned above, this corresponds to 11 / 7 = `r round(11/7, 2)` unit change on an 11-point scale.
Hence, a four-point change in media use (e.g., a complete stop) should result in a `r round(11/7, 2)`-point change in life satisfaction.
In a statistical regression, _b_ estimates the change in the dependent variable if the independent variable increases by one point.
We would therefore expect a SESOI of _b_ = `r round(11/7, 2)` / 4 = `r round(11/7*.25, 2)`.
For affect, which was measured on a 5-point scale, our SESOI would be _b_ = `r round(5 / 7, 2)` / 4 = `r round(5 / 7 * .25, 2)`.
Because we are agnostic as to whether the effects are positive or negative, the null region will include negative and positive effects.
Finally, in order not to exaggerate precision and to be less conservative, these numbers are reduced to nearby thresholds.[^rounding] 
Together, this leads to a null region ranging from _b_ = -.30 to _b_ = .30 for life satisfaction, and _b_ = -.15 to _b_ = .15 for positive and negative affect.

@@

<!-- As explained above, for well-being the null region was between _b_ = -.30 and _b_ = .30 (i.e., _b_ = |.30|). -->
To illustrate, if the 95% confidence interval falls completely within the null-region (e.g., _b_ = .20, [95% CI: .15, .25]), the hypothesis that the effect is only trivial is supported.
If the effects interval and the null region overlap (e.g., _b_ = .30, [95% CI: .25, .35]), the hypothesis is not supported and the results are considered inconclusive, while a meaningful negative effect is rejected.
If the confidence falls completely outside of the null-region (e.g., _b_ = .40, [95% CI: .35, .45]), the hypothesis is rejected and the existence of a meaningful positive effect is supported.

<!-- Responses were individually checked for patterns such as straight-lining or missing of inverted items. -->
<!-- X clear cases were removed. -->
<!-- Respondents who answered less than 50% of all questions were removed. -->
<!-- The remaining missing responses were imputed using predictive mean matching. -->

@@

The hypothesis was analyzed using mixed effects models, namely random effect within-between models (REWB)[@bellFixedRandomEffects2019]. 
Three models were run, one for each dependent variable.
The data were hierarchical, and responses were separately nested in participants and waves (put differently, participants and waves were implemented as random effects).
Nesting in participants allowed to separate between-person relations from within-person effects.
Nesting in waves allowed to control for general exogenous developments, such as general decrease in well-being in the population, for example due to lockdown measures.
Thus, there was no need additionally to control for specific phases or measures of the lockdown.
Predictors were modeled as fixed effects.
They included all types and channels of social media communication separated into within and between-person factors, as well as stable and varying covariates.
All predictors were included simultaneously and in each of the three models.

The factorial validity of the scales were tested with confirmatory factor analyses (CFA).
Because Mardia’s test showed that the assumption of multivariate normality was violated, I used the more robust Satorra-Bentler scaled and mean-adjusted test statistic (MLM) as estimator.
To avoid overfitting, I tested the scales on more liberal fit criteria (CFI > .90, TLI > .90, RMSEA <. .10, SRMR < .10) [@klinePrinciplesPracticeStructural2016].
Finally, REWB-models cannot model scales as latent variables.
To increase precision, I therefore exported factor scores from the CFAs for positive and negative affect.

For more information on the analyses, for example a complete documentation of the models and their results, see [companion website](https://tdienlin.github.io/Austrian_Corona_Panel/index.html).

@@

To capture positive affect, respondents were asked how often in the last week they felt (a) calm and relaxed, (b) happy, and (c) full of energy.
The response options were 1 (_never_), 2 (_on some days_), 3 (_several times per week_), 4 (_almost every day_), and 5 (_daily_).
The scale showed good factorial fit, `r fit_txt(cfa_aff_pos)`.
Reliability was also very high, $\omega$ = `r rel_aff_pos %>% my_round("std")`.

>>>>

abstract          : "In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop 'doomscrolling'. In this preregistered study, I investigate whether using social media for COVID-19 related reasons might put personal well-being at risk. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects within between models, controlling for several stable and varying confounders Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. The findings suggest that fears that social media use during times of crisis impairs well-being are likely to be unfounded."
keywords          : "COVID-19, Coronavirus, well-being, affect, life satisfaction, social media use, news use, communication, random effects within between model, panel study, longitudinal."

@@

During the COVID-19 pandemic, 
numerous events unfolded in quick succession.
Several open questions emerged.
How dangerous is the virus? 
Is it spreading in my region? 
How is it transmitted, and how can I protect myself?
Because for many it was (and still is) a matter of life or death, people aimed to stay informed regarding the latest developments.
Governments around the world implemented safety measures, such as wearing masks, keeping physical distance, or enforcing lockdowns.
In this extraordinary situation, many people used media excessively to attain information, and especially social media were at an all time high [@statistaAverageDailyTime2021].

Some people could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed "doomscrolling" emerged:
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children [@kleinDarklySoothingCompulsion2021].
As doomscrolling increased it became doubtful whether such social media is helpful, or whether it created an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
These concerns seem justified: 
A study with 6,233 people from Germany that was conducted during the pandemic found that "[f]requency, duration and diversity of media exposure were positively associated with more symptoms of depression" [@bendauAssociationsCOVID19Related2021, p. 283].

@@

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated communication [@meierComputermediatedCommunicationSocial2020a].
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
<!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
Well-being can be differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018].
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being, because it represents a general appraisal of one's life.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020].
<!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a].
Whereas the first four levels focus on the _channel_, the last two address the _type_ of communication.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together provides a nuanced understanding of communication.

First, I investigate how different types of communication affect well-being.
Specifically, I differentiate between active and passive use.
I distinguish (a) _reading_ (passive), _posting_ (active), and _liking and sharing_ COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects. 
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. 
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about _general_ social media use during times of COVID, but on social media use _focused_ on COVID-19 related content. 
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

@@

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
Together, the literature emphasizes potentially negative effects of social media as news use [see also @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

The question of whether and how social media use affects well-being _in general_, on the other hand, is well-researched.
This also holds true for the different types of communication such as active or passive use.
A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small [@meierComputermediatedCommunicationSocial2020a]---potentially too small to matter.
What determines whether or not an effect is considered small or trivial?
As a starting point, we could refer to standardized effect sizes.
According to @cohenPowerPrimer1992, small effect sizes start at _r_ = .10.
And indeed, several if not most of the current meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @fergusonThisMetaanalysisScreen2021].

These overviews are well aligned with several individual studies employing advanced methods [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020].
For example, @beyensSocialMediaUse2021 reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
In conclusion, most effects are likely somewhere between trivial and small.
I therefore expect that also in the case of COVID-19 related social media use effects will be trivial to small.

@@

First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which is more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news can be depressing, perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects.
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

@@

Together, the strongest argument to me is that _in general_ the effects of social media on well-being are, on average, small at best.
Because this study only looks at _one part_ of social media use---namely, COVID-19 related interactions---it is very focused, diminishing the overall potential of the effects even further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->
Therefore, I expect that COVID-19 related communication on social media does not affect well-being in a meaningful or relevant way.

@@

Testing this hypothesis, however, is not trivial.
First, in contrast to most hypotheses typically posited in the social sciences it implicitly contains an effect size, a so-called smallest effect size of interest (SESOI).
Effectively testing this hypothesis necessitates defining what's considered a "trivial effect size" and what's not.
Above I already referred to standardized effect sizes.
However, standardized effect sizes should only be a first step toward evaluating an effect's relevance [@baguleyStandardizedSimpleEffect2009].
Standardized effect sizes are determined by a sample's variance,[^cohensd] which is problematic: 
The question of whether or not social media use affects a particular person in a relevant way should not depend on the variance in the sample in which that person's data were collected.
Instead, it should depend on absolute criteria.

@@

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. 
Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to @normanInterpretationChangesHealthrelated2003, people can reliably distinguish _seven_ levels of satisfaction with health.
So if satisfaction is measured on a 7-point scale, we would state that a four unit change in social media use should result in a one unit change in life satisfaction. 
(For more information, see Method section "Inference criteria.")

@@

To illustrate, consider the following example.
Imagine that a person suddenly starts using social media much more than usual, and then after some time  becomes less satisfied with their life.
Eventually, use and life satisfaction return to prior levels.
If this happens to several people at the same time, in a longitudinal study we could then observe a significant effect of social media use on life satisfaction.
However, it could also be the case that during the study there was a major exogenous event (say, a pandemic), which caused large parts of the working population to loose their jobs.
Hence, the causal effect reported above was confounded, because in reality it was the pandemic that caused both social media use to rise and life satisfaction to go down.

Thus, only when controlling for _all_ relevant confounders, we can correctly estimate causality without bias [@rohrerThinkingClearlyCorrelations2018].
Obviously, we can never be entirely sure to have included all confounders, which makes absolute statements regarding causality virtually impossible.
In addition, when determining the overall causal effect, we need to make sure _not_ to control for mediating variables [@rohrerThinkingClearlyCorrelations2018], for doing so would bias our assessment of the causal effect.
Complicating matters further, it is often unclear if a variable is a mediator or a confounder.[^collider]
However, despite all these caveats, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly.
The aim should therefore be to collect as many varying and nonvarying confounders as possible (which it seems is seldom done in our field), while knowing that absolute certainty regarding causality cannot be reached.

@@

When searching for suitable candidates for confounders, we should look out for variables that affect both media use and well-being. 
Controlling for these factors isolates the actual effect of social media use on well-being.
We can also control for variables that affect only social media use or well-being. 
However, in doing so not much is gained or lost, because the effects of social media use would remain virtually the same [@klinePrinciplesPracticeStructural2016; but see @mcelreathYesterdayClass2021].

In this study, I hence plan to control for the following variables, which either have already been shown to affect both social media use and well-being or which are likely to do so, and which also are not mediators:
<!-- (I'll additionally include variables that likely affect only well-being, also to obtain a comparison benchmark for social media effects): -->
gender, age, education, Austria country of birth, Austria country of birth of parents, text-based news consumption, video-based news consumption, residency Vienna, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, satisfaction with democracy, disposition to take risks, and locus of control.
I will not control for variables such as trust in institutions or trust in media, because these variables might be influenced by social media use to a meaningful extent.

@@

Finally, one precondition of causality is temporal order. 
The cause needs to precede the effect.
Finding the right interval between cause and effect is crucial.
For example, if we want to measure the effect of alcohol consumption on driving performance, it makes a big difference if driving performance is measured one minute, one hour, one day, or one week after consumption.
<!-- Finding the right interval is difficult. -->
If variables are stable, longer intervals are needed; if they fluctuate, shorter intervals.
In the case of well-being, we need shorter intervals for affect and longer ones for life satisfaction.
Still, choosing the right interval is challenging, because especially short intervals are hard to implement in practice and often require advanced methods such as experience sampling (also known as in situ measurement or ambulant assessment) [@schnauber-stockmannMobileDevicesTools2020].

@@

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, p-value) were preregistered on the Open Science Framework. 
The (anonymous) preregistration can be accessed here: https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3.
Because in this study I analyzed data from an already existing large-scale data set, all of these steps were done prior to accessing the data.
The preregistration was designed on the basis of the panel documentation online [@kittelAustrianCoronaPanel2020].
In some cases I could not execute the analyses as I had originally planned, for example because some properties of the variables only became apparent when inspecting the actual data.
The most relevant deviations are reported below, and a complete list of all changes can be found in the online [companion website](https://tdienlin.github.io/Austrian_Corona_Panel/index.html) (https://tdienlin.github.io/Austrian_Corona_Panel/index.html).

@@

Because the data were analyzed post-hoc, no a-priori sample size planning on the basis of power analyses was conducted.
The sample is large, and it is hence well-equipped reliably to detect also small effects.
<!-- , which is why no exact post hoc power analysis were conducted. -->
In addition, because such large samples easily generate significant _p_-values even for very small effects, it helps that the hypotheses were tested with a smallest effect size of interest-approach.
To this end, I adopted the interval testing approach as proposed by @dienesUsingBayesGet2014.
On the basis of the SESOI, I then defined a null region.
In what follows, I explain how I determined the SESOI and the null region.

In this study, life satisfaction was measured on an 11-point scale.
If people can reliably differentiate 7 levels as mentioned above, this corresponds to 11 / 7 = `r round(11/7, 2)` unit change on an 11-point scale.
Hence, a four-point change in media use (e.g., a complete stop) should result in a `r round(11/7, 2)`-point change in life satisfaction.
In a statistical regression analysis, _b_ estimates the change in the dependent variable if the independent variable increases by one point.
We would therefore expect a SESOI of _b_ = `r round(11/7, 2)` / 4 = `r round(11/7*.25, 2)`.
For affect, which was measured on a 5-point scale, our SESOI would be _b_ = `r round(5 / 7, 2)` / 4 = `r round(5 / 7 * .25, 2)`.
Because we're agnostic as to whether the effects are positive or negative, the null region includes negative and positive effects.
Finally, in order not to exaggerate precision and to be less conservative, these numbers are reduced to nearby thresholds.[^rounding] 
Together, this leads to a null region ranging from _b_ = -.30 to _b_ = .30 for life satisfaction, and _b_ = -.15 to _b_ = .15 for positive and negative affect.

@@

<!-- As explained above, for well-being the null region was between _b_ = -.30 and _b_ = .30 (i.e., _b_ = |.30|). -->
To illustrate what this means in practice, if the 95% confidence interval falls completely within the null-region (e.g., _b_ = .20, [95% CI: .15, .25]), the hypothesis that the effect is trivial is supported.
If the effects interval and the null region overlap (e.g., _b_ = .30, [95% CI: .25, .35]), the hypothesis is not supported and the results are considered inconclusive, while a meaningful negative effect is rejected.
If the confidence falls completely outside of the null-region (e.g., _b_ = .40, [95% CI: .35, .45]), the hypothesis is rejected and the existence of a meaningful positive effect is supported.

<!-- Responses were individually checked for patterns such as straight-lining or missing of inverted items. -->
<!-- X clear cases were removed. -->

@@

The hypothesis was analyzed using mixed effects models, namely random effect within-between models (REWB)[@bellFixedRandomEffects2019]. 
Three models were run, one for each dependent variable.
The data were hierarchical, and responses were separately nested in participants and waves (i.e., participants and waves were implemented as random effects).
Nesting in participants allowed to separate between-person relations from within-person effects.
Nesting in waves allowed to control for general exogenous developments, such as general decreases in well-being in the population, for example due to lockdown measures.
Thus, there was no need additionally to control for specific phases or measures of the lockdown.
Predictors were modeled as fixed effects.
They included social media communication types and channels, separated into within and between-person factors, as well as stable and varying covariates.
All predictors were included simultaneously and in each of the three models.

The factorial validity of the scales were tested with confirmatory factor analyses (CFA).
Because Mardia’s test showed that the assumption of multivariate normality was violated, I used the more robust Satorra-Bentler scaled and mean-adjusted test statistic (MLM) as estimator.
To avoid overfitting, I tested the scales on more liberal fit criteria (CFI > .90, TLI > .90, RMSEA <. .10, SRMR < .10) [@klinePrinciplesPracticeStructural2016].
Finally, REWB-models cannot model latent variables.
To increase precision, I therefore exported factor scores from the CFAs for positive and negative affect.
Respondents who answered less than 50% of all questions were removed.
The remaining missing responses were imputed using predictive mean matching.

For more information on the analyses, a complete documentation of the models and results, see [companion website](https://tdienlin.github.io/Austrian_Corona_Panel/index.html).

@@

To capture positive affect, respondents were asked how often in the last week they felt (a) calm and relaxed, (b) happy, and (c) full of energy.
The response options were 1 (_never_), 2 (_on some days_), 3 (_several times per week_), 4 (_almost every day_), and 5 (_daily_).
The scale showed good factorial fit, `r fit_txt(cfa_aff_pos)`.
Reliability was high, $\omega$ = `r rel_aff_pos %>% my_round("std")`.

For negative affect, respondents were asked how often in the last week they felt (a) lonely, (b) aggravated, (c) so depressed, that nothing could lift you up, (d) very nervous, (e) anxious, and (h) glum and sad.
The response options were 1 (_never_), 2 (_on some days_), 3 (_several times per week_), 4 (_almost every day_), and 5 (_daily_).
The scale showed good factorial fit, `r fit_txt(cfa_aff_neg)`.
Reliability was high, $\omega$ = `r rel_aff_neg %>% my_round("std")`.
